---
title: "Predictive_PS4(Q1 except d)_735"
author: "Srinjana Sen"
date: "2026-03-29"
output: word_document
---

# PREDICTIVE ANALYTICS
## Problem Set 4
### 1. Problem to demonstrate the fitting of a binary response variable using logistic regression
a. Consider the Weekly data available in the ISLR package of R. Use the data from 1990 to 2008 as train data and the data on 2009-2010 as test data.\ 
```{r}
library(ISLR)
data(Weekly)
str(Weekly)

train=Weekly[Weekly$Year<=2008,]
test =Weekly[Weekly$Year>2008,]

nrow(train)
nrow(test)
```

b. Using the training data,fit a logistic regression with Direction as the response and the five lag variables plus Volume as predictors. Are there any significant predictors? If not,which of the tests is reporting the least p value. Interpret the coefficient corresponding to that test.\ 
```{r}
log_reg=glm(Direction ~ Lag1+Lag2+Lag3+Lag4+Lag5+Volume,
            data=train,
            family=binomial)

summary(log_reg)
```

```{r}
coef(log_reg)["Lag2"]
```


c. Estimate the odds of market going “up” at median values of the predictors.\ 
```{r}
median=apply(train[,c("Lag1","Lag2","Lag3","Lag4","Lag5","Volume")],
                 2,median)

median_df=as.data.frame(t(median))

prob=predict(log_reg,newdata=median_df,type="response")

odds=prob/(1-prob)

prob
odds
```

e. Using the fitted model and the predictor values from the test data,find an optimal cut point that minimizes TPR(1−FPR). Report the confusion matrix corresponding to the optimal cut-point value. From the confusion matrix compute the test error rate.\
```{r}
test_prob=predict(log_reg,newdata=test,type="response")

library(pROC)

roc_obj=roc(test$Direction,test_prob)

tpr=roc_obj$sensitivities
fpr=1-roc_obj$specificities
threshold=roc_obj$threshold

metric=tpr*(1-fpr)

opt_index=which.max(metric)
opt_cutoff=threshold[opt_index]

opt_cutoff

test_pred=ifelse(test_prob>opt_cutoff,"Up","Down")

table(Predicted=test_pred,Actual=test$Direction)

mean(test_pred!=test$Direction)
```

f. Is there any change in the test error rate if we have worked with only the two predictors lag 1 and lag 2?\ 
```{r}
red_fit=glm(Direction ~ Lag1+Lag2,
                data=train,
                family=binomial)

summary(red_fit)

red_test_prob=predict(red_fit,newdata=test,type="response")

red_roc_obj=roc(test$Direction,red_test_prob)

red_tpr=red_roc_obj$sensitivities
red_fpr=1-red_roc_obj$specificities
red_threshold=red_roc_obj$threshold

red_metric=red_tpr*(1-red_fpr)
red_opt_cutoff=red_threshold[which.max(red_metric)]

red_test_pred=ifelse(red_test_prob>red_opt_cutoff,"Up","Down")

table(Predicted=red_test_pred,Actual=test$Direction)

mean(red_test_pred != test$Direction)
```

